Moment Constraints Make Learning Latent-Variable Models Easier

### 2014-01-31 doc/icml2014/directed.tex
Updated undirected section
Updated draft

The tensor factorization method attacks the heart of the non-convexity
  in latent variable models.
The technique is central in solving several other problems with more
  complex correlation structure, for example latent Dirichlet allocation
  \citep{anandkumar12lda}, mixture of Gaussians \citep{hsu13spherical},
  etc.\findcite{more things}.

Once we recover the conditional moments $O\oft{i,j} \eqdef \Pr(x_i
  | h_j)$ for some $i,j$, we present a systematic approach to learn the
  conditional probability tables for every clique for a class of directed
  graphical models.
We begin by illustrating our approach with an example, followed by
  describing the simple conditions for which the approach works and some
  theoretical properties.

@@

\begin{figure}
  \centering
  \input{figures/grid.tikz}
  \caption{A directed grid model.}
  \label{fig:grid}
\end{figure}

Consider the directed grid model shown in \figureref{grid} which
  captures all the different dependency structures possible in
  a Bayesian network and is hence a good example for intuition\reword.
The model has eight observed variables $x^a_1, x^b_1 \cdots, x^a_4, x^b_4$ and four
  hidden variables $h_1, \cdots, h_4$.
The parameters of this model are the conditional probability tables
  $\pi \eqdef \Pr(h_1), T \eqdef \Pr(h_2 | h_1) = \Pr(h_3 | h_1),
  V \eqdef \Pr(h_4 | h_2, h_3)$ and $O \eqdef \Pr(x^a_i | h_i)
  =  \Pr(x^b_i | h_i)$. 
Finally, we assume $O$ and $T$ have full column rank.

\paragraph{Estimating $~O$}

Note that the observed variables $x^a_1, x^b_1, x^a_2$ are
  conditionally independent given $h_1$, allowing us to use
  \TensorFactorize, as reviewed in \sectionref{setup}, to recover
  the $O$.
We can exploit parameter sharing between the different $x$ to aggregate
  statistics from additional sets of independent
  observed variables, or {\em views}, e.g. $\{x^a_2, x^b_2, x^a_3\}$, to
  arrive at a better estimate of $O$.

%Intuitively, we expect the observations to tell us about the underlying
%  hidden state, which should make learning each factor easier. 
%For example, when $O = I$ the observed variables
%  tell us exactly which hidden state we are in; the optimal procedure to
%  estimate factors is simply counting the co-occurrences of the hidden
%  states.

\paragraph{Estimating $~\pi$}

The moments of $x^a_1$,$\mO_1 \eqdef \Pr(x^a_1)$ are directly related to
  $\pi$; $\mO_1 = O \pi$. 
By \assumptionref{full-rank}, $O$ has full column rank and thus can be
  inverted to recover $\pi$; $\pi = \pinv O \mO_1$.

\paragraph{Estimating $~T$}

Similarly, we can write down the moments of $x^a_1, x^a_2$, $\mO_{12}
  \eqdef \Pr(x^a_1, x^a_2)$ in terms of the pairwise marginals $\mH_{12}
  \eqdef \Pr(h_1, h_2)$ and solve for $\mH_{12}$;
\begin{align*}
  \mO_{12} &= O \mH_{12} O^T \\
  \mH_{12} &= \pinv O \mO_{12} \pinvt O.
\end{align*}

$T$ can be recovered from the $\mH_{12}$ by suitable normalization.

\paragraph{Estimating $V$}

Finally, we can estimate $V$ from the third-order moments $\mO_{234} \eqdef \Pr(x^a_2, x^a_3, x^a_4)$ as
follows,
\begin{align*}
  \mO_{234} &= \mH_{234}(O, O, O) \\
  \mH_{234} &= \mO_{234}(\pinv O, \pinv O, \pinv O).
\end{align*}

@@

Fundamentally, our algorithm takes advantage of having a view for every
  hidden variable in a clique to help identify its hidden state. The
  ideal property we wish our graphical models to possess is the
  following:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
For every clique $\sC$ of $\sG$ and every hidden variable $h \in \sC$,
  there is some observed variable, $x_{v_h}$, which is conditionally
  independent of the rest of the clique given $h$ and whose conditional
  moment $\mOpp{v_h}{h} \eqdef \Pr(x_{v_h} | h)$ is known. 
We call $x_{v_h}$ the exclusive view of $h$ in $\sC$.
\end{property}

Before we identify a family of models which entail this property, let us
  show that it is sufficient to learn the parameters of $\sG$.

Consider any clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$. Let $x_{v_j}$ be the exclusive
  view for $h_{i_j}$ in $\sC$ and $\sV = \{x_{v_1}, \cdots, x_{v_m}\}$
  be the set of views for the clique $\sC$.
We can write down the moments $\mO_\sV$,
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  &= \sum_{h_{i_1} = 1}^{d} \cdots \sum_{h_{i_1} = 1}^{d} 
      \Pr(\Sh{\sC}) \Pr(\Sx{\sV} \given \Sh{\sC}) \\
      &= \sum_{h_{i_1} = 1}^{d} \cdots \sum_{h_{i_1} = 1}^{d} \Pr(\Sh{\sC}) 
          \Pr(x_{i_1} | h_{v_1}) \cdots \Pr(x_{i_m} | h_{v_m}) \\
    &= Z_{\sC}(\mOpp{v_1}{i_1},\cdots,\mOpp{v_m}{i_m}).
\end{align*}

If each $\mOpp{v_j}{i_j}$ has full column rank, then we can recover the
hidden marginals $Z_\sC$ by inverting the conditional moments,
\begin{align*}
  Z_{\sC} &= \mO_\sV(\pinv{\mOpp{v_1}{i_1}},\cdots,\pinv{\mOpp{v_m}{i_m}}).
\end{align*}
Finally, the conditional probability tables for $\sC$ can easily be gotten from
  $Z_\sC$ via conditioning.

For each $\mOpp{v_j}{i_j}$ to have full column rank, it is equivalent
  to the following generalized assumption of
  \assumptionref{full-rank},

\begin{assumption} 
  \label{asm:full-rank-plus}
  For every clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$, the
  marginal distribution $Z_\sC$ derived by the parameters is full rank
  for every mode-unfolding of $Z_\sC$.
\end{assumption}

Intuitively, this assumption says that the marginal probability of every state for any hidden variable is non-zero.

\begin{lemma}
  \label{lem:goodness}

  Given \assumptionref{full-rank-plus}, (a) the marginal distribution
  $\Pr(h_i) \succ 0$ for every $h_i$ in the graph and (b)
  $\mOpp{v_j}{i_j}$ has full column rank.
\end{lemma}
\begin{proof}
  By variable elimination.
\end{proof}

\paragraph{Required structural properties}

We will now provide a sufficient condition for
\propertyref{exclusive-views} to hold. 
Recall that given three observed
variables, $x_1, x_2, x_3$ that are independent conditioned on $h$,
\TensorFactorize allows us to recover the conditional moments $\mOp{1},
\mOp{2}, \mOp{3}$. We define a hidden variable that has this capability to be a bottleneck.
\begin{definition}(Bottleneck)
  A hidden variable $h$ is said to be a bottleneck if there exists at
    least three conditionally independent observed variables, or views,
    $x_1, x_2, x_3$. 
  Let $\xB{h}$ be the set of views for $h$.
\end{definition}

The only property a graphical models must have is the following,
\begin{property}
  \label{prop:bottleneck}
  Every hidden variable is a bottleneck.
\end{property}

We note that \propertyref{bottleneck} can be relaxed if some cliques
  share parameters;
For example, because we can identify the observation potential from just
  $x^a_1, x^b_1$ and $x^a_2$, we do not require that $h_2, h_3$ or $h_4$
  be bottlenecks, and can correspondingly omit the observations $x^b_2,
  x^b_3$ and $x^b_4$.
In this case, we only require that the hidden variables in distinct
  cliques share parameters.

It is not immediately obvious that \propertyref{bottleneck} implies
  \propertyref{exclusive-views}. 
Another consideration is that while
  \TensorFactorize gives us the conditional moments for every
  bottleneck, it can not recover the conditional probability table when
  there are multiple parents for an observed variable.
The following lemma shows us how to handle this case.

\begin{lemma}(Canonical form) 
  Every directed graphical model can be transformed into one in which
  the observed variables are leaves with exactly one parent.
\end{lemma}
\begin{proof}
  \begin{figure}
    \label{fig:reduction}
    \centering
    \subimport{figures/}{reduction.tikz}
    \caption{Reduction to canonical form.}
  \end{figure}

@@

Applying \TensorFactorize to every bottleneck gives us candidate views
  for every hidden variable. Now we show that every clique also has
  exclusive views.

\begin{lemma}(Bottlenecks guarantee exclusive views)
  Provided \propertyref{bottleneck} holds, every hidden variable
    $h_{i_1}, \cdots, h_{i_m}$ in a clique $\sC$ has a exclusive view.
\end{lemma}
\begin{proof}
  Let $h$ be a hidden variable in $\sC$, let $\sV_{h}$ be the set of
    views for it and let $\sC^- \eqdef \sC \setminus \{h\}$.
  By definition, the views of $h$ are conditionally independent, viz.
    every d-connecting path between two views $x_1, x_2 \in \sV_h$ must pass through
    $h$.
  A view $x \in \sV_{h}$ is exclusive for $h$ iff $x \not\in \sV_{h'} ~ \forall h'
  \in \sC^-$

  Suppose $h$ had no unique view, then $\sV_{h} \subseteq \Union_{h' \in \sC^-} \sV_{h'}$. 
  However, this would imply that for every $x_1, x_2 \in \sV_{h}$, there
    exists some $h_1, h_2 \in \sC^-$ (not necessarily distinct) such that
    $x_1$ is d-connected to $h_1$ and $x_2$ is d-connected $h_2$.
  By participation in the clique, either $h_1 = h_2$ or $h_1$ is
    d-connected to $h_2$, implying that $x_1$ and $x_2$ are d-connected
    via a path that does not pass through $h$, contradicting the
    statement that $x_1, x_2 \in \sV$.

@@

\paragraph{Aligning the views}
\todo{How do we make sure all the recovered $O$ are align? Shouldn't be Hard}

With this lemma in place, we present the full algorithm, \LearnMarginals
in \algorithmref{directed}.

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnMarginals}
  \label{algo:directed}
  \begin{algorithmic}
    \REQUIRE A graphical model $\sG$ satisfying \textbf{(P1)}, \textbf{(P2)}, data $\sD$
    \ENSURE Marginals $Z_\sC$ for every clique $\sC \in \sG$

      \FOR{$h_i \in H$} 
        \STATE Apply \TensorFactorize to learn conditional moments
        $\mOpp{v}{i}$ for every $x_v \in \xB h$.

%    \COMMENT{Recover observation potentials $O$ using bottlenecks}
      \ENDFOR
%      \COMMENT{\textbf{Step 2:} Recover clique potentials from the piecewise likelihood.}
\FOR{every clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$} 
\STATE Identify unique views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
\STATE $\mH_\sC \gets \mO_{\Sx{\sV}}( \pinv{\mOpp{v_1}{i_i}}, \cdots, \pinv{\mOpp{v_m}{i_m}} )$.
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

\paragraph{Theoretical properties}

As each step of \LearnMarginals is a consistent estimator, it is
  straightforward that as whole it is a consistent estimator for $\sG$.
We describe its asymptotic complexity to see what the dependence on the
  parameters of the problem are.

\begin{lemma}(Sample complexity for $\sC$)

We require some number of samples to learn the factor for $\sC$.
\end{lemma}
\begin{proof}

For the sake of simplicity, let us assume that our dataset of $n$
  samples is partitioned into two sets.
  The first $\bytwo{n}$ samples are used to estimate $\mOpphat{v}{i}$ for
  every $h_i \in H$ and $x_v \in \sV_{h}$ in step 1 of \algorithmref{directed}.
\citet{anandkumar13tensor} show that 
\begin{align*}
  \|\mOpphat{v}{i} - \mOpp{v}{i}\|_F &= O( \frac{k \pi_{\max}^2}{\sigma_k(M_2)^5} ). 
\end{align*}

The next $\bytwo{n}$ samples are used to compute the moments $M_\sV$ for
  every clique in step 2.

We have for each clique,
\begin{align*}
  Z_{\sC} &= \mO_\sV(\pinv{\mOpp{v_1}{i_1}},\cdots,\pinv{\mOpp{v_m}{i_m}}) \\
\end{align*}

Choosing to represent $Z_\sC$ as a vector and $\mOpp{\cdot}{\cdot}$ as a matrix,
let $\pinv{\mOpp{\sV}{\sC}} \eqdef \pinv{\mOpp{v_1}{i_1}} \otimes \cdots \otimes \pinv{\mOpp{v_m}{i_m}}$,
\begin{align*}
  Z_{\sC} &= \pinv{\mOpp{\sV}{\sC}} \mO_\sV.
\end{align*}

We know the asymptotic distribution of $M_\sV$ goes as,
\begin{align*}
  \sqrt{n}(\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV),
\end{align*}
where $\Sigma_\sV$ is the variance. Because $M_\sV$ describes the discrete marginal distribution, we have 
\begin{align*}
  \Sigma_\sV &= \diag  M_\sV  - M_\sV M_\sV^T,
\end{align*}

Thus, $Z_\sC$ goes as
\begin{align*}
  \sqrt{n}(\hat Z_\sV - Z_\sV) \convind \sN(0, \pinvt{\mOpp{\sV}{\sC}} \Sigma_\sV \pinv{\mOpp{\sV}{\sC}}).
\end{align*}

\todo{Convert from $Z_\sC$ to the actual CPTs}.

\end{proof}

>>>>

The tensor factorization method attacks the heart of the non-convexity
  in latent variable models.
  Once we recover the conditional moments $\mOpp{v}{j} \eqdef \Pr(x_v
  | h_j)$ for some $v, j$, we present a systematic approach to learn the
  conditional probability tables for every clique for a class of directed
  graphical models.
We begin by illustrating our approach with an example, followed by
  describing the algorithm in full generality.
For ease of exposition, we will describe our algorithm entirely in the
  context of directed graphical models, though it generalizes to factor
  graphs.
We also present the algorithm solely in terms of the method of moments; in
  \sectionref{piecewise}, we show how we can get a statistically more
  efficient estimator using convex optimization.

@@

\begin{figure}
  \centering
  \input{figures/grid.tikz}
  \caption{A directed grid model.\todo{Annotate features}}
  \label{fig:grid}
\end{figure}

Consider the directed grid model shown in \figureref{grid} which
  captures all the different dependency structures possible in
  a Bayesian network.
The model has eight observed variables $x^a_1, x^b_1 \cdots, x^a_4, x^b_4$ and four
  hidden variables $h_1, \cdots, h_4$.
The parameters of this model are the conditional probability tables
  $\pi \eqdef \Pr(h_1), T \eqdef \Pr(h_2 | h_1) = \Pr(h_3 | h_1),
  V \eqdef \Pr(h_4 | h_2, h_3)$ and $O \eqdef \Pr(x^a_i | h_i)
  =  \Pr(x^b_i | h_i)$. 
Finally, we assume $O$ and $T$ have full column rank.

\paragraph{Estimating $O$}
Note that the observed variables $x^a_1, x^b_1, x^a_2$ are
  conditionally independent given $h_1$, allowing us to use
  \TensorFactorize, as reviewed in \sectionref{setup}, to recover
  the $O$.

\paragraph{Estimating $\pi$}
The moments of $x^a_1$,$\mO_1 \eqdef \Pr(x^a_1)$ are directly related to
  $\pi$; $\mO_1 = O \pi$. 
By \assumptionref{full-rank}, $O$ has full column rank and thus can be
  inverted to recover $\pi$; $\pi = \pinv O \mO_1$.

\paragraph{Estimating $T$}
Similarly, we can write down the moments of $x^a_1, x^a_2$, $\mO_{12}
  \eqdef \Pr(x^a_1, x^a_2)$ in terms of the pairwise marginals $\mH_{12}
  \eqdef \Pr(h_1, h_2)$ and solve for $\mH_{12}$;
\begin{align*}
  \mO_{12} &= O \mH_{12} O^T \\
  \mH_{12} &= \pinv O \mO_{12} \pinvt O.
\end{align*}
$T$ can be recovered from the $\mH_{12}$ by suitable normalization.

\paragraph{Estimating $V$}
Finally, we can estimate $V$ from the third-order moments $\mO_{234} \eqdef \Pr(x^a_2, x^a_3, x^a_4)$ as
follows,
\begin{align*}
  \mO_{234} &= \mH_{234}(O, O, O) \\
  \mH_{234} &= \mO_{234}(\pinv O, \pinv O, \pinv O).
\end{align*}

@@

Fundamentally, our algorithm takes advantage of having a view for every
  hidden variable in a clique to help identify its hidden state. We define this condition as follows,
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
A clique $\sC$ of $\sG$ is said to posses exclusive views if for every hidden variable $h_i \in \sC$,
  there is some observed variable, $x_{v}$, which is conditionally
  independent of the rest of the clique given $h_i$ and whose conditional
  moment $\mOpp{v}{i} \eqdef \Pr(x_{v} | h_i)$ is known. 
We call $x_v$ the exclusive view of $h_i$ in $\sC$.
\end{property}

\paragraph{Learning cliques}

We now show this condition is sufficient to learn the marginal
  distribution of the clique.
Consider any clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$ for which this property holds. Let
  $x_{v_j}$ be the exclusive view for $h_{i_j}$ in $\sC$ and $\sV
  = \{x_{v_1}, \cdots, x_{v_m}\}$ be the set of views for the clique
  $\sC$.
We can write down the moments $\mO_\sV$,
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  &= \sum_{\vh \in \sH}
      \Pr(\Sh{\sC}) \Pr(\Sx{\sV} \given \Sh{\sC}) \\
      &= \sum_{\vh \in sH} \Pr(\Sh{\sC}) 
          \Pr(x_{i_1} | h_{v_1}) \cdots \Pr(x_{i_m} | h_{v_m}) \\
    &= Z_{\sC}(\mOpp{v_1}{i_1},\cdots,\mOpp{v_m}{i_m}).
\end{align*}

If each $\mOpp{v_j}{i_j}$ has full column rank, then we can recover the
hidden marginals $Z_\sC$ by inverting the conditional moments,
\begin{align*}
  Z_{\sC} &= \mO_\sV(\pinv{\mOpp{v_1}{i_1}},\cdots,\pinv{\mOpp{v_m}{i_m}}).
\end{align*}
Finally, the conditional probability tables for $\sC$ can easily be gotten from
  $Z_\sC$ via conditioning.
\algorithmref{learnclique} summarizes the procedure, \LearnClique.

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnClique}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE A $\sC$ with \propertyref{exclusive-views}.
    \ENSURE The marginal distribution of the clique, $Z_\sC$.
      \STATE Identify unique views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE $\mH_\sC \gets \mO_{\Sx{\sV}}( \pinv{\mOpp{v_1}{i_i}}, \cdots, \pinv{\mOpp{v_m}{i_m}} )$.
  \end{algorithmic}
\end{algorithm}

\paragraph{Structural properties}

If we can guarantee that every clique in a graph possesses exclusive
  views, then we can apply \LearnClique iteratively to learn the entire
  model.
In the example, we were able to guarantee \propertyref{exclusive-views}
  by identifying a hidden variable, $h^1$, with three conditionally
  independent observed variables, $x^a_1, x^b_1, x^a_2$, followed by
  using \TensorFactorize to learn the conditional moments.
We define such a hidden variable to be a bottleneck;
\begin{definition}(Bottleneck)
  A hidden variable $h$ is said to be a bottleneck if there exists at
    least three conditionally independent observed variables, or views,
    $x_1, x_2, x_3$. 
  Let $\sV_{h}$ be the set of views for $h$.
\end{definition}
Finally, the claim we make is that we can recover parameters for any
  graphical model where every hidden variable is a bottleneck;
\begin{property}
  \label{prop:bottleneck}
  Every hidden variable is a bottleneck.
\end{property}
We note that \propertyref{bottleneck} can be relaxed if some cliques
  share parameters;
For example, because we can identify the observation potential from just
  $x^a_1, x^b_1$ and $x^a_2$, we do not require that $h_2, h_3$ or $h_4$
  be bottlenecks, and can correspondingly omit the observations $x^b_2,
  x^b_3$ and $x^b_4$.
In this case, we only require that the hidden variables in distinct
  cliques share parameters.

Applying \TensorFactorize to every bottleneck gives us candidate views
  for every hidden variable. 
To proceed, we need to show two things; (i) under what assumptions
  does \assumptionref{full-rank} hold for each bottleneck and (ii)
  that the candidate views actually imply that \propertyref{exclusive-views}
  holds for every clique in the graph.

To address the first, we will need that the following assumption on the
  parameters of the model:
\begin{assumption} 
  \label{asm:full-rank-plus}
  For every clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$, let
    $Z_\sC$ be the marginal distribution for that clique. 
  Then every mode-$i$ unfolding of $Z_\sC$ has full column rank.
\end{assumption}
Intuitively, the assumption guarantees that the marginal probability of
  any hidden state is non-zero.

Before we establish the second condition, we describe a reduction to
  a canonical form that will allow us to handle the case where an observed
  variable has multiple parents.

\begin{lemma}(Canonical form) 
  \label{lem:reduction}
  Every directed graphical model can be transformed into one in which
  the observed variables are leaves with exactly one parent.
\end{lemma}
\begin{proof}
  \begin{figure}
    \label{fig:reduction}
    \centering
    \subimport{figures/}{reduction.tikz}
    \caption{Reduction to canonical form.}
  \end{figure}

@@

\begin{lemma}(Bottlenecks guarantee exclusive views)
  Provided \propertyref{bottleneck} holds, every hidden variable
    $h_{i_1}, \cdots, h_{i_m}$ in a clique $\sC$ has a exclusive view.
\end{lemma}
\begin{proof}
  By \lemmaref{reduction}, we can assume w.l.o.g. that every observed
  variable is a leaf in the graph, causing any v-structures in the graph
  to be active. This allows us to reason about independence simply by
  checking whether or not two nodes are connected or not; note that this
  proof applies for undirected graphs as well.

  Let $h$ be a hidden variable in $\sC$, let $\sV_{h}$ be the set of
    views for it and let $\sC^- \eqdef \sC \setminus \{h\}$.
  By definition, the views of $h$ are conditionally independent, viz.
    every connecting path between two views $x_1, x_2 \in \sV_h$ must pass through
    $h$.
  A view $x \in \sV_{h}$ is exclusive for $h$ iff $x \not\in \sV_{h'} ~ \forall h'
  \in \sC^-$

  Suppose $h$ had no unique view, then $\sV_{h} \subseteq \Union_{h' \in \sC^-} \sV_{h'}$. 
  However, this would imply that for every $x_1, x_2 \in \sV_{h}$, there
    exists some $h_1, h_2 \in \sC^-$ (not necessarily distinct) such that
    $x_1$ is connected to $h_1$ and $x_2$ is connected $h_2$.
  By participation in the clique, either $h_1 = h_2$ or $h_1$ is
    connected to $h_2$, implying that $x_1$ and $x_2$ are connected
    via a path that does not pass through $h$, contradicting the
    statement that $x_1, x_2 \in \sV$.

@@

With this lemma in place, we present the full algorithm, \LearnMarginals~
in \algorithmref{directed}.

\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\begin{algorithm}
  \caption{\LearnMarginals}
  \label{algo:directed}
  \begin{algorithmic}
    \REQUIRE A graphical model $\sG$ satisfying \propertyref{bottleneck}, data $\sD$
    \ENSURE Marginals $Z_\sC$ for every clique $\sC \in \sG$

      \FOR{$h_i \in H$} 
        \STATE Apply \TensorFactorize to learn conditional moments
        $\mOpp{v}{i}$ for every $\sV_h$.

%    \COMMENT{Recover observation potentials $O$ using bottlenecks}
      \ENDFOR
%      \COMMENT{\textbf{Step 2:} Recover clique potentials from the piecewise likelihood.}
\FOR{every clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$} 
  \STATE Apply \LearnClique to learn the marginals $\mH_\sC$.
\ENDFOR
  \end{algorithmic}
\end{algorithm}

\paragraph{Consistency and Sample Complexity}

\LearnMarginals~ combines two consistent algorithms, \TensorFactorize and
  \LearnClique, and is thus consistent itself. Using results from
  \citet{anandkumar12moments,anandkumar13tensor} we can show that that
  learning $\mOpp{v_1}{i}$ for the bottleneck $h_i$ with views $x_{v_1},
  x_{v_2}, x_{v_3}$ has the following sample complexity,
\begin{align*}
  \|\mOpphat{v_1}{i} - \mOpp{v_1}{i}\|^2_F &= \frac{1}{\sqrt{n}} O( \frac{k {\pi\oft{i}}_{\max}^2}{\sigma_k(M_{v_1,v_2})^5} ). 
\end{align*}

We next describe the asymptotic variance for \LearnClique when using the
  method of moments.
\begin{lemma}(Asymptotic variance of $Z_\sC$)
  \label{lem:mom-variance}  
  The asymptotic variance of $Z_\sC$, $\Sigma^\mom_\sC$ is,
  \begin{align*}
    \Sigma^{(\mom)} &= \mOppit{\sV}{\sC} \Sigma_\sV \mOppi{\sV}{\sC},
  \end{align*}
  where $\pinv{\mOpp{\sV}{\sC}} \eqdef \pinv{\mOpp{v_1}{i_1}} \otimes
    \cdots \otimes \pinv{\mOpp{v_m}{i_m}}$, a $k^m \times d^m$ matrix, and 
\begin{align*}
  \Sigma_\sV &= \diag  M_\sV  - M_\sV M_\sV^T,
\end{align*}
  is the variance of the observations $M_\sV$.
\end{lemma}
\begin{proof}
We have for each clique,
\begin{align*}
  Z_{\sC} &= \mO_\sV(\mOppi{v_1}{i_1},\cdots,\mOppi{v_m}{i_m}) \\
\end{align*}

Choosing to represent $Z_\sC$ and $M_\sV$ as vectors and
$\mOppi{v_1}{i_1} \otimes \cdots \otimes \mOppi{v_m}{i_m}$ as the
matrix, we can rewrite this representation as,
\begin{align*}
  Z_{\sC} &= \pinv{\mOpp{\sV}{\sC}} \mO_\sV.
\end{align*}

We know the asymptotic distribution of $M_\sV$ goes as,
\begin{align*}
  \sqrt{n}(\hat M_\sV - M_\sV) \convind \sN(0, \Sigma_\sV),
\end{align*}
where $\Sigma_\sV$ is the variance of the observations. 
Finally, we have that $Z_\sC$ goes as
\begin{align*}
  \sqrt{n}(\hat Z_\sC - Z_\sC) \convind \sN(0, \pinvt{\mOpp{\sV}{\sC}} \Sigma_\sV \pinv{\mOpp{\sV}{\sC}}).
\end{align*}
\end{proof}

Note that extending these results to finite sample bounds can be done
  via a straightforward application of perturbation bounds.

### 2014-01-31 doc/icml2014/intro.tex
Updated draft

% 1. Latent variable models are good.
Latent variable models offer a succinct representation of a rich model
family. 
% 2. Learning them is hard.
Despite their success across many fields
\cite{quattoni04crf,haghighi06prototype,liang06discrimative,kirkpatrick10painless},
learning these models remains a difficult problem due to the
non-convexity of the likelihood. Local optimization (e.g.
expectation-maximization) is the standard approach, but is susceptible
to local optima.

% 3. People have approached unsupervised learning with the MoM magic sauce, but the sauce is limited.
Recently, unsupervised learning techniques based on the method of moments and
spectral decomposition have offered a refreshing and promising perspective on
this learning problem \citep{hsu09spectral,anandkumar11tree,anandkumar12moments,anandkumar12lda,hsu12identifiability,balle11transducer,balle12automata}.
These methods exploit the linear algebraic properties of the model and
factorize the moments into parameters, providing strong theoretical guarantees.
However, these methods are not as universally applicable as EM, applying to a limited set of models.

@@

% 4. State what we do: exploit moment constraints to make the problem easier.
In this work, we exploit the spectral method to learn moment constraints
on the observed variables and show how these constraints lead to consistent parameter estimates for a broad class of graphical models.

% 5. We get moments from third-order tensors from bottlenecks and factorize them into marginals.
Our approach is illustrated in \figureref{approach}. 
First, we identify conditionally independent observed variables $x_1,
  x_2, x_3$ for each hidden variable $h$ (bottlenecks). 
We use the tensor factorization algorithm of
  \citet{anandkumar12moments,anandkumar13tensor} to identify the
  conditional moments.
Given these factors, we show that for every clique of hidden variables
  $\sC$, the piecewise likelihood of the observed variables is convex in
  the marginal distribution. 
This guarantees that we can recover the true marginal distributions
  using EM; finally, with some post-processing, we are able to
  consistently recover the true parameters of the model.
In \sectionref{piecewise}, we detail our algorithm and describe the
  settings and technical assumptions in which this approach is guaranteed
  to work.
As a particular example, we are able to recover parameters for the
  grid model depicted in \figureref{approach} under the condition that
  $O$ is full-rank.

For undirected log-linear latent variable models, this approach allows us to 
  recover the marginals for each clique, given which the learning
  problem is convex (\sectionref{undirected}).
Finally, our approach gracefully extends to more general model families
  in which we are unable to recover the marginal distributions for all
  cliques; in such scenarios, we use posterior regularization to constrain
  EM. 
Empirically, we find that this approach leads to better solutions
  (\sectionref{experiments}).

>>>>

% 1. Latent variable models are good.
Latent variable graphical models offer a succinct representation of
a rich model family. 
% 2. Learning them is hard.
Despite their success across many fields
\cite{quattoni04crf,beaumont2004bayesian,haghighi06prototype,liang06discrimative,kirkpatrick10painless},
learning these models remains a difficult problem due to the
non-convexity of the likelihood. Local optimization (e.g.
expectation-maximization) is the standard approach, but is susceptible
to local optima.

% 3. People have approached unsupervised learning with the MoM magic sauce, but the sauce is limited.
Recently, unsupervised learning techniques based on the method of moments and
spectral decomposition have offered a refreshing and promising perspective on
this learning problem \citep{hsu09spectral,anandkumar11tree,anandkumar12moments,anandkumar12lda,hsu12identifiability,balle11transducer,balle12automata}.
These methods exploit the linear algebraic properties of the model and
factorize the moments into parameters, providing strong theoretical guarantees.
However, these methods apply to a limited set of models, and are thus
not as universally applicable as EM.

@@

% 4. State what we do: exploit moment constraints to make the problem easier.
In this work, we exploit the spectral method to learn moment constraints
  on the observed variables and show how these constraints lead to consistent parameter estimates 
  for any graphical model with the property that there exist at least
  three ``views'' or conditionally independent observed variables for
  every hidden variable (\propertyref{bottleneck}).

% 5. We get moments from third-order tensors from bottlenecks and factorize them into marginals.
Our approach is illustrated in \figureref{approach}. 
First, we identify the views $x_1, x_2, x_3$ for each hidden variable $h$. 
We then use the tensor factorization algorithm of
  \citet{anandkumar12moments,anandkumar13tensor} to identify the
  conditional moments.
Given these factors, we show that there exists a set of observed
  variables for every clique such that the likelihood is convex in the
  parameters of that clique. 
This guarantees that we can recover the true marginal distributions
  using EM; finally, with some post-processing, we are able to
  consistently recover the true parameters of the model.
In \sectionref{directed}, we detail our algorithm and describe the
  settings and technical assumptions in which this approach is guaranteed
  to work.
As a particular example, we are able to recover parameters for the
  grid model depicted in \figureref{approach} under the condition that
  $O$ and $T$ is full-rank.

Our approach also applies to undirected log-linear latent variable
  models, where we use the recovered marginals to make the learning
  problem convex (\sectionref{undirected}).
Finally, our approach gracefully extends to more general model families
  in which we are unable to recover the marginal distributions for all
  cliques; in such scenarios, we use posterior regularization to constrain
  EM.

### 2014-01-31 doc/icml2014/related.tex
Updated draft

\section{Related work}

Anandkumar - linear bayesian nets

Le song - trees

David sontag - noisy or

>>>>

\section{Related work}
\label{sec:related}

Parameter estimation for latent variable models using the method of
  moments has received much attention in last few years.
Our work uses the \TensorFactorize algorithm presented by
  \citet{anandkumar13tensor}, though \citet{anandkumar12moments} first
  studied the consistent parameter estimation algorithm for the
  three-view mixture model.
%\todo{What about Mossel and Roch?}
Several authors have extended this work to settings that do not meet the
  criterion \propertyref{bottleneck}, but rely on knowing more about
  the underlying correlation structure of the problem.
\citet{anandkumar12lda} are able to recover parameters for correlated
  discrete mixtures, e.g. latent Dirichlet allocation; their approach
  follows that of the three-view mixture model, with a pre-processing step
  that subtracts the correlation between the views.
\citet{anandkumar12linear} study a special family of Bayesian networks
  in which the observed variables are linearly related to a sparse set of
  hidden variables. Once again, the key step is to identify the
  correlations between observed variables and exclude them from the views.
\citet{halpern13noisyor} propose an algorithm to recover factors for
  a more densely connected bipartite noisy-or network; they also use
  bottlenecks to learn parameters, but exploit properties of the model
  to negate correlations from learned edges, allowing for more
  bottlenecks to be discovered. 
In comparison, our work applies to a general family of models, including
  latent tree models, directed and undirected grids, with no assumptions
  on the correlation structure aside from independences encoded in the
  graph.

Another prominent line of work in the method of moments community has
  focussed on recovering parameters for a transformation that admits an {\em observable
  representation}, i.e. one whose marginals can be directly computed from the data.
\citet{song2011spectral} propose an algorithm to learn an observable
  operator representations for latent tree graphical models, like the
  one in \figureref{examples-tree}, assuming \propertyref{bottleneck}. 
Their approach is similar to ours, but restricted to trees with observed
  variables at the leaves.
\citet{parikh12spectral} extend this approach to general graphical
  models in which \propertyref{bottleneck} holds, but consider the latent
  junction tree representation. 
Consequently, the observable representations have size atleast that of
  a node in the junction tree, which will be of the order of the treewidth
  of the graph. 
Our algorithm only constructs moments of the order of the cliques in the
  graph, which can be exponentially smaller. 
For example, an $n\times n$ grid model has a treewidth of $n$, but each
  clique is of size at most $2$ in the undirected case and $3$ in the
  directed case.
In some sense, an observable operator re-formulation of our algorithm
  subsumes this work.
